\newsection{Decisions}

\begin{center}
\textbf{Note: all code snippets from here on are assumed to be in a main method} \\ (they will not compile if not).
\end{center}

\subsection{Motivation}
We now know basic data types and how to use them. Now, we will see how to execute conditional code based on values. We will cover \verb|if|, \verb|if-else|, and \verb|switch| statements.

\subsection{If Statements}
Suppose we were given a task of printing ``True" only when a value entered by the user was larger than another value. Currently, we do not know of a way to do this. Therefore, we have \verb|if| statements. They have the form:
\begin{center}
$if(condition) \{ ... \}$
\end{center}
where ``..." means other lines of code, and condition is evaluated to a \verb|boolean| value. The rule for \verb|if| statements is: if the condition evaluates to \verb|true|, then the code between the braces is executed; if \verb|false|, then it is not executed. If we have one line of code that will be executed in the \verb|if| statement, then the braces are optional:
\begin{lstlisting}
int x = 8;
if (x > 7)
     System.out.print("This one line");
\end{lstlisting}
However, the second line after the \verb|if| statement in the following example will not execute as if it were in the \verb|if| statement, even though it is indented:
\begin{lstlisting}
int x = 8;
if (x > 7)
     System.out.print("This one line");
     System.out.println("What about this line?");
\end{lstlisting}
The above code is functionally equivalent to:
\begin{lstlisting}
int x = 8;
if (x > 7) {
     System.out.print("This one line");
}
System.out.println("What about this line?"); // always prints
\end{lstlisting}
To fix this, we add braces:
\begin{lstlisting}
int x = 8;
if (x > 7) {
     System.out.print("This one line");
     System.out.println("What about this line?");
}
\end{lstlisting}
and therefore, both lines will print if the condition evaluates to \verb|true| (which it does in this case).

\par Now suppose we want to execute some code when the condition is \verb|true|, and some other code when the condition is \verb|false| (and only when it is \verb|false|). Therefore, we need an \verb|if-else|statement, which is of the form:
\begin{center}
$if(condition) \{...\} else \{...\}$
\end{center}
If the condition is \verb|false|, then the code in the \verb|else| section will automatically execute, and the code in the \verb|if| section will not. If the condition is \verb|true|, then the \verb|else| section will not execute, and the \verb|if| section will.

\par Suppose we have been given a task of printing ``true" if the user's input is greater than or equal to 8, and ``false" otherwise. Our code then will look like:
\begin{lstlisting}
Scanner s = new Scanner(System.in);
int input = s.nextInt();
if (input >= 8) {
     System.out.print("true");
} else {
     System.out.print("false");
}
\end{lstlisting}
We can even chain multiple \verb|if-else| statements together (and even nested ones!):
\begin{lstlisting}
Scanner s = new Scanner(System.in);
int input = s.nextInt();
if (input >= 8) {
     if (input == 8) {
          System.out.print("Equal to 8");
     } else {
          System.out.print("Larger than 8");
     }
} else if (input >= 4) {
     if (input == 4) {
          System.out.print("Equal to 4");
     } else {
          System.out.print("Larger than 4, < 8");
     }
} else {
     System.out.print("Less than 4");
}
\end{lstlisting}

\subsection{Switch Statements}
An alternative to \verb|if-else| statements are switch statements, because \verb|if-else| statements can be very complex and lengthy. A \verb|switch| statement structure consists of 1 or more \verb|case| statements (with an optional \verb|default| case). The programmer supplies a \verb|switch| statement with a value:
\begin{lstlisting}
switch(value) {
     // ...
}
\end{lstlisting}
where \verb|value| is a primitive type or a \verb|String| (Note: \verb|String| is only allowed in later versions of Java). For each of the case statements, \verb|value| is compared to the value supplied in the \verb|case| statement (see form below). Only if they are equal will the code after the \verb|case| statement be executed.
\begin{itemize}
\item Each \verb|case| statement is of the form: \verb|case value: /* statements */|. At the end of the code section for a \verb|case| statement, an optional \verb|break;| statement is allowed. However, not putting this statement at the end of the code section for the case statement will immediately execute the next \verb|case| statement, regardless of the value.
\item An optional \verb|default:| case statement at the end of the \verb|switch| block is allowed, which is executed if the input value is not equal to any other \verb|case| statement's value.
\end{itemize}
For example, if we were to print ``1" if the input value is equal to 1, and ``2" if equal to 2, and ignore any other inputs, our code might look like:
\begin{lstlisting}
Scanner s = new Scanner(System.in);
int input = s.nextInt();
switch(input) {
     case 1:
          System.out.print("1");
          break;
     case 2:
          System.out.print("2");
          break;
}
\end{lstlisting}
Now, if we remove the \verb|break| statements:
\begin{lstlisting}
Scanner s = new Scanner(System.in);
int input = s.nextInt();
switch(input) {
     case 1:
          System.out.print("1");
     case 2:
          System.out.print("2");
}
\end{lstlisting}
this code will output ``12" if the input is 1, and ``2" if the input is 2, which is not the behavior we might want. This is because we removed the \verb|break| statements, and so the code flows through until it hits a \verb|break| statement. What a \verb|break| statement does is exit the \verb|switch| statement altogether.

\par On the other hand, if we want to add the requirement of printing ``other" if we have any other value than 1 or 2, then we use a \verb|default| statement:
\begin{lstlisting}
Scanner s = new Scanner(System.in);
int input = s.nextInt();
switch(input) {
     case 1:
          System.out.print("1");
          break;
     case 2:
          System.out.print("2");
          break;
     default:
          System.out.print("other");
          break;
}
\end{lstlisting}
We can see that this code can be constructed in an \verb|if-else| statement fashion. However, sometimes \verb|switch| and \verb|case| statements are easier to understand and read.

\subsection{Written Exercises}

\begin{enumerate}

\item What is the output of the following code?
\begin{lstlisting}
int depth = 30;
if (depth >= 29) {
     System.out.print("Bigger than 8!");
     System.out.print("Don't swim!");
}
System.out.println("Yes, you can swim.");
\end{lstlisting}

\item What is the output of the following code?
\begin{lstlisting}
int mystery1 = 12;
int mystery2 = 42;
System.out.print("You have: ");
if (mystery1 >= 8)
     System.out.print("1 ");
if (mystery2 <= 50 && mystery1 <= 12)
     System.out.print("2 ");
System.out.println("3.");
\end{lstlisting}

\item If k holds a value of the type \verb|int|, then the value of the expression:
\begin{lstlisting}
k <= 10 || k > 10
\end{lstlisting}
\begin{enumerate}
\item[a)] must be true
\item[b)] must be false
\item[c)] could be either true or false
\item[d)] is a value of type \verb|int|
\end{enumerate}

\item For the following code, fill in the missing condition to check if \verb|str1| and \verb|str2| are the same.
\begin{lstlisting}
String str1 = "Java is fun";
String str2 = "Java is fun";
if ( /* */ )
     System.out.println("String1 and String2 are the same");
else
     System.out.println("String1 and String2 are different");
\end{lstlisting}

\item Evaluate the following expressions, assuming that \verb|x = -2| and \verb|y = 3|.
\begin{lstlisting} 
x <= y 
(x < 0) || (y < 0)
(x <= y) && (x < 0)
((x + y) > 0) && !(y > 0)
\end{lstlisting}

\item Write the output of the following code:
\begin{lstlisting}
int grade = 45;
if (grade >= 70)
     System.out.println("passing");
if (grade < 70)
     System.out.println("dubious");
if (grade < 60)
     System.out.println("failing");
\end{lstlisting}

\item Write the output of the following code:
\begin{lstlisting}
String option = "A";
if (option.equals("A"))
     System.out.println("addRecord");
if (option.compareTo("A") == 0)
     System.out.println("deleteRecord");
\end{lstlisting}

\item Write the output of the following code:
\begin{lstlisting}
double x = -1.5;
if (x < -1.0)
     System.out.println("true");
else
     System.out.println("false");
     System.out.println("after if...else");
\end{lstlisting}

\item Write the output of the following code:
\begin{lstlisting}
int j = 8;
double x = -1.5;
if (x >= j)
     System.out.println("x is high");
else
     System.out.println("x is low");
\end{lstlisting}

\item Write the output of the following code:
\begin{lstlisting}
double x = -1.5;
if (x <= 0.0) {
     if (x < 0.0)
          System.out.println("neg");
     else
          System.out.println("zero");
}
else
     System.out.println("pos");
\end{lstlisting}

\end{enumerate}

\subsection{Programming Exercises}

\begin{enumerate}

\item Write a program that asks for 3 integers and prints the median value of the 3 integers, using only \verb|if| statements.

\item Write code that ensures that an \verb|int| variable called \verb|number| is an odd integer.

\end{enumerate}